Inferring cosmology from gravitational waves using non-parametric detector-frame mass distribution

### 2024-10-29 src/tex/ms.tex
[showyourwork] overleaf sync

\abstract{
    The challenge of understanding the Universe's dynamics, particularly the Hubble tension, requires precise measurements of the Hubble constant.
    Building upon the existing spectral-siren method, which capitalizes on population information from gravitational-wave sources, this paper explores an alternative way to analyze the population data to obtain the cosmological parameters in $\Lambda$CDM.
    We demonstrated how non-parametric methods, which are flexible models that can be used to agnostically reconstruct arbitrary probability densities, can be incorporated into this framework and leverage the detector-frame mass distribution to infer the cosmological parameters.
    We tested our method with mock data and applied it to the real data from the LIGO-Virgo-KAGRA Collaboration.
}

@@

In this example, instead of transforming the source-frame population model to the detector frame for each $H_0$ and performing a grid search, we define a function that calculates $d_\mathrm{JS}$ for a given set of the free parameters.
We then use the optimization algorithm mentioned in Sec.~\ref{sec:remapping} to find the optimized free parameters that minimize $d_\mathrm{JS}$.
The optimization method requires bounds for the free parameters, which we show in Table~\ref{tab:bounds}.
\begin{table}[htbp]
    \caption{Free parameter bounds for the joint inference presented in Section~\ref{sec:inference_multi}.}
    \begin{tabular}{cc}
        \toprule
        Parameter & Bounds \\
        \midrule
        $H_0$ & $[10, 300]$ \\
        $\alpha$ & $[1.01, 10]$ \\
        $\mu$ & $[10, 70]$ \\
        $\sigma$ & $[0.01, 10]$ \\
        \botrule
    \end{tabular}
    \label{tab:bounds}
\end{table}
The initial guesses for the optimization are chosen randomly from a uniform distribution within the bounds.
The result of the remapping is shown in Fig.~\ref{fig:simulation_result_multi}.
\begin{figure}
    \script{plot_simulation_result_multi.py}
    \includegraphics[width=\linewidth]{figures/simulation_result_multi.pdf}
    \caption{
        Optimised parameter distribution for the mock data inference. Simulated values are marked in red.
    }
    \label{fig:simulation_result_multi}
\end{figure}
The recovered distribution, whose uncertainty budget comes from the remapping of the individual-\ac{PE} uncertainty in the non-parametric reconstruction, is consistent with the simulated values.

@@

In this section, we apply our method to the real data from the \ac{LVK}.
We included all $70$ events labeled as BBH in Table~I of \cite{KAGRA:2021duu}, which are different from the events used in \cite{LIGOScientific:2021aug}.

@@

We then perform the remaining steps similar to Sec.~\ref{sec:inference_multi}.
We set the bounds for $\kappa$ to $[-10, 10]$, and the bounds for other free parameters are the same as in Table~\ref{tab:bounds}.
The result of the remapping is shown in Fig.~\ref{fig:real_result_multi}, and it is consistent with the posterior distribution reported in \cite{KAGRA:2021duu}.
\begin{figure}
    \script{plot_real_result_multi.py}
    \includegraphics[width=\linewidth]{figures/real_result_multi.pdf}
    \caption{
        Optimised parameter distribution for the real data inference.
    }
    \label{fig:real_result_multi}
\end{figure}

@@

\bibliography{bib}

>>>>

\abstract{
    The challenge of understanding the Universe's dynamics, particularly the Hubble tension, requires precise measurements of the Hubble constant.
    Building upon the existing spectral-siren method, which capitalizes on population information from gravitational-wave sources, this paper explores an alternative way to analyze the population data to obtain the cosmological parameters in $\Lambda$CDM.
    We demonstrated how non-parametric methods, which are flexible models that can be used to agnostically reconstruct arbitrary probability densities, can be incorporated into this framework and leverage the detector-frame mass distribution to infer the cosmological parameters.
    We tested our method with mock data and applied it to $70$ real events from the third gravitational-wave transient catalog of the LIGO-Virgo-KAGRA Collaboration.
}

@@

In this example, instead of transforming the source-frame population model to the detector frame for each $H_0$ and performing a grid search, we define a function that calculates $d_\mathrm{JS}$ for a given set of the free parameters.
We then use the optimization algorithm mentioned in Sec.~\ref{sec:remapping} to find the optimized free parameters that minimize $d_\mathrm{JS}$.
The optimization method requires bounds for the free parameters, which we show in Table~\ref{tab:bounds}.
\begin{table}[htbp]
    \caption{Free parameter bounds for the joint inference presented in Section~\ref{sec:inference_multi}.}
    \begin{tabular}{cc}
        \toprule
        Parameter & Bounds \\
        \midrule
        $H_0$ & $[10, 300]$ \\
        $\alpha$ & $[1.01, 10]$ \\
        $\mu$ & $[10, 70]$ \\
        $\sigma$ & $[0.01, 10]$ \\
        \botrule
    \end{tabular}
    \label{tab:bounds}
\end{table}
The initial guesses for the optimization are chosen randomly from a uniform distribution within the bounds.
The result of the remapping is shown in Fig.~\ref{fig:simulation_result_multi}.
\begin{figure}
    \script{plot_simulation_result_multi.py}
    \includegraphics[width=\linewidth]{figures/simulation_result_multi.pdf}
    \caption{
        Optimized parameter distribution for the mock data inference. Simulated values are marked in red.
    }
    \label{fig:simulation_result_multi}
\end{figure}
The recovered distribution, whose uncertainty budget comes from the remapping of the individual-\ac{PE} uncertainty in the non-parametric reconstruction, is consistent with the simulated values.

@@

In this section, we apply our method to the real data from the \ac{LVK}.
We included all $70$ events labeled as BBH in Table~I of \cite{KAGRA:2021duu}, these correspond to all the events with $\mathrm{FAR_{min}} < 1\ \mathrm{yr}^{-1}$, where $\mathrm{FAR_{min}}$ is the smallest false alarm rate over all search pipelines.
These events are different from the events used in \cite{LIGOScientific:2021aug}, therefore, the results are not directly comparable.

@@

We then perform the remaining steps similar to Sec.~\ref{sec:inference_multi}.
We set the bounds for $\kappa$ to $[-10, 10]$, and the bounds for other free parameters are the same as in Table~\ref{tab:bounds}.
The result of the remapping is shown in Fig.~\ref{fig:real_result_multi}, and it is consistent with the posterior distribution reported in \cite{KAGRA:2021duu}.
\begin{figure}
    \script{plot_real_result_multi.py}
    \includegraphics[width=\linewidth]{figures/real_result_multi.pdf}
    \caption{
        Optimized parameter distribution for the real data inference.
    }
    \label{fig:real_result_multi}
\end{figure}
The result implies that \output{real_H0.txt}, which has an uncertainty much smaller than the one reported in \cite{KAGRA:2021duu}.
This is because the uncertainty only includes the uncertainty from the reconstruction, which underestimates the true uncertainty which also includes the mismatch between the population model and the observed population distribution.

@@

\bibliography{bib.bib}
